Informing IPCC accounting of forest carbon using the global forest carbon database (ForC v4.0)

### 2023-05-16 doc/manuscript/ESSDmanuscript.Rmd
add ref, incorporate small ForC updates

Aboveground biomass, which is typically <200 t C ha^-1^ but can exceed 700 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by the IPCC as "all biomass of living vegetation above the soil including stems, stumps, branches, bark, seeds, and foliage" [@ipcc_good_2003;@ipcc_2006_2006].
IPCC's guidance is that the understory may be excluded the understory if it constitutes a "minor" component, *where quantitative definitions of "understory" and "minor" are not provided*<!--verify (issue 40)-->, but where a commonly applied minimum size sampling threshold for mature forests would be 10 cm stem diameter at breast height (DBH).
A recent study characterizing the contributions of trees in different DBH classes to ecosystem C stocks and fluxes found that trees 1 - 10 cm DBH contributed up to ~8% aboveground biomass, ~17% aboveground woody net primary productivity ($ANPP_{woody.stem}$), and ~20% woody mortality ($M_{woody}$) of mature closed-canopy forests worldwide [@piponiot_distribution_2022].
In regrowth forests, woodlands, or savannas, small trees and shrubs contribute a much larger proportion of C stocks and fluxes [@lutz_global_2018; @piponiot_distribution_2022; @hughes_biomass_1999], and, correspondingly, biomass estimates for these ecosystems tend include smaller size classes. 
While IPCC guidance specifies that all living vegetation should be included in biomass estimates, forest censuses and biomass estimates do not consistently include life forms other than dicot trees (e.g., lianas, ferns, palms, bamboo), although these do tend to be censused when they constitute a large proportion of the biomass [@refs_from_ForC].
Further, it is important to note that the IPCC definition of aboveground biomass excludes standing dead wood, which is included in remote sensing biomass estimates [@duncanson_aboveground_2021].

>>>>

Aboveground biomass, which is typically <200 t C ha^-1^ but can exceed 700 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by the IPCC as "all biomass of living vegetation above the soil including stems, stumps, branches, bark, seeds, and foliage" [@ipcc_good_2003;@ipcc_2006_2006].
IPCC's guidance is that the understory may be excluded the understory if it constitutes a "minor" component, *where quantitative definitions of "understory" and "minor" are not provided*<!--verify (issue 40)-->, but where a commonly applied minimum size sampling threshold for mature forests would be 10 cm stem diameter at breast height (DBH).
A recent study characterizing the contributions of trees in different DBH classes to ecosystem C stocks and fluxes found that trees 1 - 10 cm DBH contributed up to ~8% aboveground biomass, ~17% aboveground woody net primary productivity ($ANPP_{woody.stem}$), and ~20% woody mortality ($M_{woody}$) of mature closed-canopy forests worldwide [@piponiot_distribution_2022].
In regrowth forests, woodlands, or savannas, small trees and shrubs contribute a much larger proportion of C stocks and fluxes [@lutz_global_2018; @piponiot_distribution_2022; @hughes_biomass_1999], and, correspondingly, biomass estimates for these ecosystems tend include smaller size classes. 
While IPCC guidance specifies that all living vegetation should be included in biomass estimates, forest censuses and biomass estimates do not consistently include life forms other than dicot trees (e.g., lianas, ferns, palms, bamboo), although these do tend to be censused when they constitute a large proportion of the biomass [e.g., @fukushima_recovery_2007].
Further, it is important to note that the IPCC definition of aboveground biomass excludes standing dead wood, which is included in remote sensing biomass estimates [@duncanson_aboveground_2021].

### 2023-05-16 doc/manuscript/ESSDmanuscript.tex
add ref, incorporate small ForC updates

\textbf{Abstract.} Forests are critical for climate change mitigation
and constitute a substantial portion of planned emissions reductions
under the 2015 Paris Agreement. Yet, the efficacy of greenhouse gas
mitigation planning and reporting is dependent upon the quality of
available emission factors data, including forest carbon (C) stocks and
changes therein. Tens of thousands of relevant forest C estimates have
been published, yet are not readily accesible to the practitioners
compiling national greenhouse gas inventories. Many of these data have,
however, been compiled in the Global Forest C database (ForC;
https://forc-db.github.io/) and stand to be of value to greenhouse gas
accounting if made available through the Emission Factor Database (EFDB)
of the International Panel on Climate Change (IPCC). Here, we develop
and document a process for semi-automated transfer of data from ForC
into the EFDB, assess the data available and transferred to date, and
provide recommendations for improving forest data collection, analysis,
and reporting to improve accounting of forest-sector greenhoouse gas
emissions and removals. We begin by reconciling terminology and mapping
ForC fields into EFDB. This process required some updates to the ForC
database structure, leading to the release of a new version of ForC
(v4.0; described here). As of May 15, 2023, ForC contained
\textasciitilde19323 independent records relevant to EFDB, 1438 of which
have been submitted to date. Among the data in ForC, there is
disproportionate representation of biomass (particularly aboveground)
stocks, with far fewer records for dead organic matter and soil C, and
relatively few or no records for net annual increments or C fluxes into
(gains) or out of (losses) the IPCC-defined C pools. Geographic
representation is also quite uneven, with the highest densities of
relevant records in temperate forests, and with relatively scant
representation of tropical forests in Africa and Asia. This distribution
of records is generally reflected in the subset of records that have
been submitted to EFDB to date. In the future, forest C estimates in
EFDB can be improved through targeted research to fill critical gaps,
reporting of information required by IPCC, and continued submission of
data from scientific publications to the EFDB. Given that climate change
is rapidly impacting the world's forests, timely reporting of recent
estimates will be especially critical to accurate forest C accounting.

@@

The Global Forest Carbon Database, ForC
(\url{https://forc-db.github.io/}), is the largest collection of
published estimates of forest C stocks, increments, and annual fluxes
\citep{anderson-teixeira_forc_2018, anderson-teixeira_carbon_2021}. ForC
includes data ingested from individual publications and relevant
databases, including the Global Reforestation Opportunity Assessment
(GROA) database \citep[database doi:
10.5281/zenodo.3983644]{cook-patton_mapping_2020}, the global soil
respiration database
\citep[SRDB-V5,][]{bond-lamberty_global_2010, jian_restructured_2021}.
As of May 15, 2023, ForC contained 39855 records from 10589 plots in
1535 distinct geographical areas, along with records of stand age and
disturbance history. As such, ForC is positioned to improve forest C
accounting through the transfer of data to EFDB. The purpose of this
publication is to document that process and provide recommendations for
future improvements.

@@

Aboveground biomass, which is typically \textless200 t C
ha\textsuperscript{-1} but can exceed 700 t C ha\textsuperscript{-1}
\citep{anderson-teixeira_carbon_2021}, is defined by the IPCC as ``all
biomass of living vegetation above the soil including stems, stumps,
branches, bark, seeds, and foliage''
\citep{ipcc_good_2003, ipcc_2006_2006}. IPCC's guidance is that the
understory may be excluded the understory if it constitutes a ``minor''
component, \emph{where quantitative definitions of ``understory'' and
``minor'' are not provided}, but where a commonly applied minimum size
sampling threshold for mature forests would be 10 cm stem diameter at
breast height (DBH). A recent study characterizing the contributions of
trees in different DBH classes to ecosystem C stocks and fluxes found
that trees 1 - 10 cm DBH contributed up to \textasciitilde8\%
aboveground biomass, \textasciitilde17\% aboveground woody net primary
productivity (\(ANPP_{woody.stem}\)), and \textasciitilde20\% woody
mortality (\(M_{woody}\)) of mature closed-canopy forests worldwide
\citep{piponiot_distribution_2022}. In regrowth forests, woodlands, or
savannas, small trees and shrubs contribute a much larger proportion of
C stocks and fluxes
\citep{lutz_global_2018, piponiot_distribution_2022, hughes_biomass_1999},
and, correspondingly, biomass estimates for these ecosystems tend
include smaller size classes. While IPCC guidance specifies that all
living vegetation should be included in biomass estimates, forest
censuses and biomass estimates do not consistently include life forms
other than dicot trees (e.g., lianas, ferns, palms, bamboo), although
these do tend to be censused when they constitute a large proportion of
the biomass \citep{refs_from_ForC}. Further, it is important to note
that the IPCC definition of aboveground biomass excludes standing dead
wood, which is included in remote sensing biomass estimates
\citep{duncanson_aboveground_2021}.

@@

To create structure for EFDB-relevant records, we added a total of 15
new variables to the set of named and defined variables (Fig. 1),
counting each pair of variables with units in C (ending in \texttt{\_C})
or organic matter (ending in \texttt{\_OM}) as one. The majority of
these were increment variables (n=11), adding to only one previously
defined increment variable (aboveground biomass increment,
\emph{delta.agb}). These are directly related to C stocks as previously
defined in ForC, with ``\emph{delta.}'' added in front of the variable
name. Further, we added variables capturing the belowground component of
woody mortality (\emph{woody.mortality\_root}) and the combined
aboveground and belowground components of woody mortality
(\emph{woody.mortality}). Although most of these variables lacked
records in ForC as of May 15, 2023, their addition gave the structure
such that records can be populated over time. Finally, to provide better
definition of the previously existing variable \emph{organic.layer},
which has a nebulous definition that reflects the varied definitions
adopted by original studies, we added two clearly defined variables:
\emph{litter} (relatively undecomposed plant material/ OL horizon), and
\emph{O.horizon} (entire O-horizon, including OL).

@@

Prior to releasing ForC v4.0, we executed several quality control
measures. First, we implemented a system of continuous integration using
GitHub Actions \citep[\emph{sensu}][]{kim_implementing_2022} to run some
automatic checks any time the master data files are updated, including
outlier tests and checks for completeness and naming consistency of
records across data files. Second, to improve information on geographic
coordinates, we created a field to record coordinate precision (Appendix
A), and flagged and reviewed records with suspected low precision.
Third, to identify erroneous climate data, we compared ForC climate
values to those extracted from WorldClim version \textbf{\#\#}
\citep{climwin_ref} based on site coordinates. Records deviating from
WorldClim values by more variable-specific thresholds (\textgreater5°C
for mean annual temperature, \textgreater7.5°C for mean temperatures of
the warmest and coldest months, or \textgreater1 for log(mean annual
precipitation in mm)) were flagged as requiring review prior to use in
analysis or transfer to EFDB.

@@

To transfer complete, reviewed ForC records into EFDB, we created R
scripts to restructure ForC records and populate EFDB's bulk import form
(``EFDB bulk import.xlsx''). Criteria for data transfer were that (1)
records had been checked against the original study and determined to be
complete and correct, and as originally presented, (2) the original
study presented values in tables or text, as opposed to the values
having been digitized from graphs or calculated based on related
variables, and (3) the records had not previously been sent to EFDB.
Once converted into EFDB format, the records were reviewed and then sent
to the IPCC's Technical Support Unit for inclusion in EFDB. Complete
records needed to be reviewed by the EFDB editorial board prior to
posting in the database -- a process that lags behind records transfer
and had not yet been completed for all records sent as of May 15, 2023.

@@

As of May 15, 2023, ForC (v4.0) contained 32693 independent records
(39855 total), 19323 of which were for the the 43 variables relevant to
EFDB (Fig. 1). These records were distributed across all forested
continents and ecozones, albeit unevenly (Fig. 2). The largest number of
records came from Asia, followed by North America, South America, and
Europe, with relatively few records from Africa, Australia, and Oceania
(Fig. 3c). Categorized by FAO ecozone, the greatest numbers of records
came from subtropical humid forests, temperate mountain systems, and
tropical rain forests, each with \textgreater2,000 independent records
(Fig. 3b). Boreal coniferous forests, temperate continental forests,
subtropical mountain systems, and tropical moist deciduous forests had
\textgreater1,000 independent records each, while other ecozones all had
\textless1,000 records. The most widely represented forest type was
needleleaf evergreen, followed by broadleaf deciduous and broadleaf
evergreen (Fig. 3a). In terms of stand age, the most represented age
class was 20-100 years, followed by \textless20 years and then
\textgreater100 years (Fig. 3d).

@@

As of May 15, 2023, we had reviewed or added 2283 EFDB-relevant records,
1438 records of which were sent to EFDB, and 376 of which have been
reviewed, accepted, and posted (Figs. 2-3, Table 2). The 37\%
attenuation between records reviewed and those sent to EFDB was
attributable to the presence of digitized records and records where a
variable's value had been calculated as the sum or difference of related
variables rather than presented directly in the text. The discrepancy
between the number of records sent and that posted to EFDB is primarily
attributable to the time required for the IPCC to review the records,
and it is not expected that many -- if any -- records will be rejected.

@@

The ForC database contains large numbers of records that could
potentially be useful for C accounting under IPCC guidelines. Here we
have developed a framework for transferring these records to the EFDB,
making them more accessible for forest C accounting under IPCC
guidelines \citep{ipcc_2006_2006, ipcc_2019_2019}. As of May 15, 2023,
we have submitted 1438 records to EFDB. Although this represents just
7\% of relevant records in ForC, it substantially increases the number
of forest land records in EFDB. The records transferred to EFDB and
present in ForC are very unevenly distributed across variables, regions,
and forest types (Figs. 2-3, Table 2), reflecting broader patterns in
allocation of research effort.

>>>>

\textbf{Abstract.} Forests are critical for climate change mitigation
and constitute a substantial portion of planned emissions reductions
under the 2015 Paris Agreement. Yet, the efficacy of greenhouse gas
mitigation planning and reporting is dependent upon the quality of
available emission factors data, including forest carbon (C) stocks and
changes therein. Tens of thousands of relevant forest C estimates have
been published, yet are not readily accesible to the practitioners
compiling national greenhouse gas inventories. Many of these data have,
however, been compiled in the Global Forest C database (ForC;
https://forc-db.github.io/) and stand to be of value to greenhouse gas
accounting if made available through the Emission Factor Database (EFDB)
of the International Panel on Climate Change (IPCC). Here, we develop
and document a process for semi-automated transfer of data from ForC
into the EFDB, assess the data available and transferred to date, and
provide recommendations for improving forest data collection, analysis,
and reporting to improve accounting of forest-sector greenhoouse gas
emissions and removals. We begin by reconciling terminology and mapping
ForC fields into EFDB. This process required some updates to the ForC
database structure, leading to the release of a new version of ForC
(v4.0; described here). As of May 16, 2023, ForC contained
\textasciitilde19323 independent records relevant to EFDB, 1438 of which
have been submitted to date. Among the data in ForC, there is
disproportionate representation of biomass (particularly aboveground)
stocks, with far fewer records for dead organic matter and soil C, and
relatively few or no records for net annual increments or C fluxes into
(gains) or out of (losses) the IPCC-defined C pools. Geographic
representation is also quite uneven, with the highest densities of
relevant records in temperate forests, and with relatively scant
representation of tropical forests in Africa and Asia. This distribution
of records is generally reflected in the subset of records that have
been submitted to EFDB to date. In the future, forest C estimates in
EFDB can be improved through targeted research to fill critical gaps,
reporting of information required by IPCC, and continued submission of
data from scientific publications to the EFDB. Given that climate change
is rapidly impacting the world's forests, timely reporting of recent
estimates will be especially critical to accurate forest C accounting.

@@

The Global Forest Carbon Database, ForC
(\url{https://forc-db.github.io/}), is the largest collection of
published estimates of forest C stocks, increments, and annual fluxes
\citep{anderson-teixeira_forc_2018, anderson-teixeira_carbon_2021}. ForC
includes data ingested from individual publications and relevant
databases, including the Global Reforestation Opportunity Assessment
(GROA) database \citep[database doi:
10.5281/zenodo.3983644]{cook-patton_mapping_2020}, the global soil
respiration database
\citep[SRDB-V5,][]{bond-lamberty_global_2010, jian_restructured_2021}.
As of May 16, 2023, ForC contained 39848 records from 10589 plots in
1535 distinct geographical areas, along with records of stand age and
disturbance history. As such, ForC is positioned to improve forest C
accounting through the transfer of data to EFDB. The purpose of this
publication is to document that process and provide recommendations for
future improvements.

@@

Aboveground biomass, which is typically \textless200 t C
ha\textsuperscript{-1} but can exceed 700 t C ha\textsuperscript{-1}
\citep{anderson-teixeira_carbon_2021}, is defined by the IPCC as ``all
biomass of living vegetation above the soil including stems, stumps,
branches, bark, seeds, and foliage''
\citep{ipcc_good_2003, ipcc_2006_2006}. IPCC's guidance is that the
understory may be excluded the understory if it constitutes a ``minor''
component, \emph{where quantitative definitions of ``understory'' and
``minor'' are not provided}, but where a commonly applied minimum size
sampling threshold for mature forests would be 10 cm stem diameter at
breast height (DBH). A recent study characterizing the contributions of
trees in different DBH classes to ecosystem C stocks and fluxes found
that trees 1 - 10 cm DBH contributed up to \textasciitilde8\%
aboveground biomass, \textasciitilde17\% aboveground woody net primary
productivity (\(ANPP_{woody.stem}\)), and \textasciitilde20\% woody
mortality (\(M_{woody}\)) of mature closed-canopy forests worldwide
\citep{piponiot_distribution_2022}. In regrowth forests, woodlands, or
savannas, small trees and shrubs contribute a much larger proportion of
C stocks and fluxes
\citep{lutz_global_2018, piponiot_distribution_2022, hughes_biomass_1999},
and, correspondingly, biomass estimates for these ecosystems tend
include smaller size classes. While IPCC guidance specifies that all
living vegetation should be included in biomass estimates, forest
censuses and biomass estimates do not consistently include life forms
other than dicot trees (e.g., lianas, ferns, palms, bamboo), although
these do tend to be censused when they constitute a large proportion of
the biomass \citep[e.g.,][]{fukushima_recovery_2007}. Further, it is
important to note that the IPCC definition of aboveground biomass
excludes standing dead wood, which is included in remote sensing biomass
estimates \citep{duncanson_aboveground_2021}.

@@

To create structure for EFDB-relevant records, we added a total of 15
new variables to the set of named and defined variables (Fig. 1),
counting each pair of variables with units in C (ending in \texttt{\_C})
or organic matter (ending in \texttt{\_OM}) as one. The majority of
these were increment variables (n=11), adding to only one previously
defined increment variable (aboveground biomass increment,
\emph{delta.agb}). These are directly related to C stocks as previously
defined in ForC, with ``\emph{delta.}'' added in front of the variable
name. Further, we added variables capturing the belowground component of
woody mortality (\emph{woody.mortality\_root}) and the combined
aboveground and belowground components of woody mortality
(\emph{woody.mortality}). Although most of these variables lacked
records in ForC as of May 16, 2023, their addition gave the structure
such that records can be populated over time. Finally, to provide better
definition of the previously existing variable \emph{organic.layer},
which has a nebulous definition that reflects the varied definitions
adopted by original studies, we added two clearly defined variables:
\emph{litter} (relatively undecomposed plant material/ OL horizon), and
\emph{O.horizon} (entire O-horizon, including OL).

@@

Prior to releasing ForC v4.0, we executed several quality control
measures. First, we implemented a system of continuous integration using
GitHub Actions \citep[\emph{sensu}][]{kim_implementing_2022} to run some
automatic checks any time the master data files are updated, including
outlier tests and checks for completeness and naming consistency of
records across data files. Second, to improve information on geographic
coordinates, we created a field to record coordinate precision (Appendix
A), and flagged and reviewed records with suspected low precision.
Third, to identify erroneous climate data, we compared ForC climate
values to those extracted from WorldClim version 2.1
\citep{vandepol_identifying_2016, bailey_climwin_2016} based on site
coordinates. Records deviating from WorldClim values by more
variable-specific thresholds (\textgreater5°C for mean annual
temperature, \textgreater7.5°C for mean temperatures of the warmest and
coldest months, or \textgreater1 for log(mean annual precipitation in
mm)) were flagged as requiring review prior to use in analysis or
transfer to EFDB.

@@

To transfer complete, reviewed ForC records into EFDB, we created R
scripts to restructure ForC records and populate EFDB's bulk import form
(``EFDB bulk import.xlsx''). Criteria for data transfer were that (1)
records had been checked against the original study and determined to be
complete and correct, and as originally presented, (2) the original
study presented values in tables or text, as opposed to the values
having been digitized from graphs or calculated based on related
variables, and (3) the records had not previously been sent to EFDB.
Once converted into EFDB format, the records were reviewed and then sent
to the IPCC's Technical Support Unit for inclusion in EFDB. Complete
records needed to be reviewed by the EFDB editorial board prior to
posting in the database -- a process that lags behind records transfer
and had not yet been completed for all records sent as of May 16, 2023.

@@

As of May 16, 2023, ForC (v4.0) contained 32693 independent records
(39848 total), 19323 of which were for the the 43 variables relevant to
EFDB (Fig. 1). These records were distributed across all forested
continents and ecozones, albeit unevenly (Fig. 2). The largest number of
records came from Asia, followed by North America, South America, and
Europe, with relatively few records from Africa, Australia, and Oceania
(Fig. 3c). Categorized by FAO ecozone, the greatest numbers of records
came from subtropical humid forests, temperate mountain systems, and
tropical rain forests, each with \textgreater2,000 independent records
(Fig. 3b). Boreal coniferous forests, temperate continental forests,
subtropical mountain systems, and tropical moist deciduous forests had
\textgreater1,000 independent records each, while other ecozones all had
\textless1,000 records. The most widely represented forest type was
needleleaf evergreen, followed by broadleaf deciduous and broadleaf
evergreen (Fig. 3a). In terms of stand age, the most represented age
class was 20-100 years, followed by \textless20 years and then
\textgreater100 years (Fig. 3d).

@@

As of May 16, 2023, we had reviewed or added 2283 EFDB-relevant records,
1438 records of which were sent to EFDB, and 376 of which have been
reviewed, accepted, and posted (Figs. 2-3, Table 2). The 37\%
attenuation between records reviewed and those sent to EFDB was
attributable to the presence of digitized records and records where a
variable's value had been calculated as the sum or difference of related
variables rather than presented directly in the text. The discrepancy
between the number of records sent and that posted to EFDB is primarily
attributable to the time required for the IPCC to review the records,
and it is not expected that many -- if any -- records will be rejected.

@@

The ForC database contains large numbers of records that could
potentially be useful for C accounting under IPCC guidelines. Here we
have developed a framework for transferring these records to the EFDB,
making them more accessible for forest C accounting under IPCC
guidelines \citep{ipcc_2006_2006, ipcc_2019_2019}. As of May 16, 2023,
we have submitted 1438 records to EFDB. Although this represents just
7\% of relevant records in ForC, it substantially increases the number
of forest land records in EFDB. The records transferred to EFDB and
present in ForC are very unevenly distributed across variables, regions,
and forest types (Figs. 2-3, Table 2), reflecting broader patterns in
allocation of research effort.
